---
title:  |
  | Ecological Forecasting
  | Lab 5 - Stan Model Intro
author: "Seth Romero"
date: "2023-02-20"
output: html_document
---

```{r, include = FALSE}
options(width = 200)
knitr::opts_chunk$set(include = FALSE, message = FALSE, warning = FALSE, fig.height = 6, fig.width = 10)
```

```{r}
library(rstan)
library(shinystan)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(patchwork)
library(kableExtra)

setwd("~/Documents/GitHub/CourseResources/EcologicalForecasting/assignments/")
```

```{stan, output.var = "stan1"}
// stan model using vectors
// compiled output as "stan1" using output.var =
data {
  int<lower=0> N;
  vector[N] y;
  vector[N] x1;
  vector[N] x2;
}

parameters {
  real b0;
  real b1;
  real b2;
  real<lower=0> sigma;
}

model {
  y ~ normal(b0 + b1*x1 + b2*x2, sigma);
}
```

```{r, include = TRUE, results = "hide"}
full_ts <- read.csv("../data/portal_timeseries.csv") %>%
  mutate(date = mdy(date)) %>%
  arrange(date) %>%
  mutate(i = seq(1, length(.$NDVI)), 
         date_intrvl = time_length((date - lag(date, n = 1)), unit = "days"))

# calculating t-1 values,
# removing final 10 dates
obs_ts <- full_ts %>% 
  mutate(prev_ndvi = lag(NDVI, n = 1),
         prev_rain = lag(rain, n = 1)) %>%
  filter(i <= (max(i) - 10))

ndviData <- list(N = dim(obs_ts)[1], 
                 y = obs_ts$NDVI, 
                 x1 = obs_ts$prev_ndvi, 
                 x2 = obs_ts$prev_rain)

# sampling() effectively the same as stan(), but preferable in markdown workflow
# allows us to use an object of class `stanmodel` which is created by the {stan} chunks
# (chunks are running `rstan::stan_model(model_code=the_text)` in the background)
M1stan <- sampling(stan1,
                   data = ndviData,
                   chains = 3,
                   iter = 2000,
                   warmup = 1000)
```

```{r}
n <- length(full_ts$NDVI)
PredData <- full_ts[(n-10): n,]
PredOut <- matrix(NA, length(parsM1$b0), 10)

for (p in 1:length(parsM1$b0)){
  NDVI = PredData$NDVI[1]
  for (t in 1:10){
    NDVI = parsM1$b0[p] + parsM1$b1[p] * NDVI + parsM1$b2[p] * PredData$rain[t]
    PredOut[p,t] = NDVI
  }
}

po <- as.data.frame(PredOut)

matplot(t(PredOut), type = 'l')

MeanP <- apply(PredOut, 2, mean)
Upper <- apply(PredOut, 2, quantile, prob = 0.975)
Lower <- apply(PredOut, 2, quantile, prob = 0.025)

stuff <- data.frame(matrix(ncol = 4, nrow = 10))
stuff[, 2] <- MeanP
stuff[, 3] <- Upper
stuff[, 4] <- Lower
stuff[, 1] <- seq(1:10)

colnames(stuff) <- c("month", "mean", "upper", "lower")
```

```{r, echo = FALSE}
parsM1 <- as.data.frame(rstan::extract(M1stan, c("b0", "b1", "b2", "sigma")))

# with lapply(), builds list of ggplot objects (in this case histograms)
build_hists <- function(parameters, column){
  ggplot(parameters, aes_string(x = column)) +
    geom_histogram(color = "#979494", fill = "#DFDFDF", bins = 20) +
    geom_vline(xintercept = mean(parameters[, column]), linetype = "dashed", linewidth = 0.75, color = "#273485") +
    ylab("Frequency") +
    xlab(column) +
    theme_classic()
}

# builds a multi-panel layout of ggplot objects stored in list
# currently using default gridding from `patchwork`
build_patchwork <- function(gglist){
  mp_hists = gglist[[1]]
  for (n in 2:length(gglist)) {
    mp_hists = mp_hists + gglist[[n]]
  }
  return(mp_hists)
}
```

```{r}
histsNDVI <- lapply(colnames(po), build_hists, parameters = po)
build_patchwork(histsNDVI)

```


***1. Develop a forecast for the mean NDVI for the last 10 months, assuming that the rainfall is known. Include a plot of the forecast timeseries with the mean (i.e. the mean of the mean NDVI) and 95% credible intervals for the mean as well as the observed data***

***2. How does the SD of the predicted mean change as we forecast forward in time? Does uncertainty become larger or smaller or stay the same?***

***3. Why is it important to account for covariance in parameter error? Do you think not including covariance would increase or decrease our predicted uncertainty in this example?***
