---
title:  |
  | Ecological Forecasting
  | Lab 5 - Stan Model Pt. II
author: "Seth Romero"
date: "2023-02-20"
output: html_document
---

<style type="text/css">
  body {
    font-size: 11pt;
    max-width: 800px;
    margin: auto;
  }
</style>

```{r, include = FALSE}
options(width = 200)
knitr::opts_chunk$set(include = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 8)
```

```{r, include = FALSE}
library(rstan)
library(shinystan)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(patchwork)
library(kableExtra)

setwd("~/Documents/GitHub/CourseResources/EcologicalForecasting/assignments/")
```

```{stan, output.var = "stan1", include = TRUE}
// stan model using vectors
// compiled output as "stan1" using output.var =
data {
  int<lower=0> N;
  vector[N] y;
  vector[N] x1;
  vector[N] x2;
}

parameters {
  real b0;
  real b1;
  real b2;
  real<lower=0> sigma;
}

model {
  y ~ normal(b0 + b1*x1 + b2*x2, sigma);
}

```

```{r, include = TRUE, results = "hide"}
full_ts <- read.csv("../data/portal_timeseries.csv") %>%
  mutate(date = mdy(date)) %>%
  arrange(date) %>%
  mutate(i = seq(1, length(.$NDVI)), 
         date_intrvl = time_length((date - lag(date, n = 1)), unit = "days"))

obs_ts <- full_ts %>% 
  mutate(prev_ndvi = lag(NDVI, n = 1),
         prev_rain = lag(rain, n = 1)) %>%
  filter(i <= (max(i) - 10)) %>%
  na.omit()

ndviData <- list(N = dim(obs_ts)[1], 
                 y = obs_ts$NDVI, 
                 x1 = obs_ts$prev_ndvi, 
                 x2 = obs_ts$prev_rain)

M1stan <- sampling(stan1,
                   data = ndviData,
                   chains = 3,
                   iter = 2000,
                   warmup = 1000)

parsM1 <- as.data.frame(rstan::extract(M1stan, c("b0", "b1", "b2", "sigma")))
```

<hr style="border-width:4px">

***1. Develop a forecast for the mean NDVI for the last 10 months, assuming that the rainfall is known. Include a plot of the forecast timeseries with the mean (i.e. the mean of the mean NDVI) and 95% credible intervals for the mean as well as the observed data.***

```{r, include = TRUE}
FullTsLength <- length(full_ts$NDVI)
PredData <- full_ts[(FullTsLength-10): FullTsLength,]
PredOut <- matrix(NA, length(parsM1$b0), 10)

for (p in 1:length(parsM1$b0)){
  NDVI = PredData$NDVI[1]
  for (t in 1:10){
    NDVI = parsM1$b0[p] + parsM1$b1[p] * NDVI + parsM1$b2[p] * PredData$rain[t]
    PredOut[p,t] = NDVI
  }
}

PredOut <- as.data.frame(PredOut)

# '2' argument is for 'margin' implying function is calculated on columns
# ('1' would apply to rows)
MeanP <- apply(PredOut, 2, mean)
Upper <- apply(PredOut, 2, quantile, prob = 0.975)
Lower <- apply(PredOut, 2, quantile, prob = 0.025)

PredCIntervals <- data.frame(matrix(ncol = 7, nrow = 10)) %>%
  mutate(X1 = full_ts$date[(FullTsLength-9):FullTsLength],
         X2 = MeanP,
         X3 = Upper,
         X4 = Lower,
         X5 = full_ts$NDVI[(FullTsLength-9):FullTsLength],
         X6 = Upper,
         X7 = Lower) %>%
  rename(Month = X1,
         "Pred. Mean" = X2,
         "Pred. Upper 95" = X3,
         "Pred. Lower 95" = X4,
         Observed = X5,
         ribbonU = X6,
         ribbonL = X7) %>%
  pivot_longer(c(2:5), names_to = "Parameters", values_to = "NDVI") %>%
  mutate(Parameters = factor(Parameters, levels = c("Observed", "Pred. Mean",
                                                    "Pred. Upper 95", "Pred. Lower 95")))

ggplot(PredCIntervals, 
       aes(x = Month, y = NDVI, 
           linetype = Parameters, color = Parameters, shape = Parameters)) +
  geom_line() + 
  geom_point(data = subset(PredCIntervals, 
                           Parameters == "Observed" | Parameters == "Pred. Mean")) +
  scale_shape_manual(values = c(16, 1, NA, NA)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotted")) +
  scale_color_manual(values = c("#166170", "#166170", "#616AA0", "#616AA0")) +
  geom_ribbon(aes(x = Month, ymin = ribbonL, ymax = ribbonU), 
              inherit.aes = FALSE, fill = "#616AA0", alpha = 0.2) +
  scale_y_continuous(limits = c(0.1, 0.4)) +
  scale_x_date(breaks = "2 months",
               date_labels = "%y %b") +
  theme_bw()
```

<hr style="border-width:4px">

***2. How does the SD of the predicted mean change as we forecast forward in time? Does uncertainty become larger or smaller or stay the same?***

```{r, include = TRUE, fig.width = 7}

SDP <- apply(PredOut, 2, sd)
PredSDdf <- as.data.frame(matrix(ncol = 2, nrow = 10)) %>%
  mutate(X1 = full_ts$date[(FullTsLength-9):FullTsLength],
         X2 = SDP) %>%
  rename(Month = X1,
         SD = X2)

ggplot(PredSDdf, aes(x = Month, y = SD)) +
  geom_line() +
  scale_x_date(breaks = "2 months",
               date_labels = "%y %b") +
  theme_bw() +
  labs(y = "SD of NDVI predicted mean")

```


***3. Why is it important to account for covariance in parameter error? Do you think not including covariance would increase or decrease our predicted uncertainty in this example?***
